Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-18 ind_files/01_Introduction.tex
Add some references

Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled.

>>>>

Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's cotinuum-of-alleles model~\citep{kimura_stochastic_1965} and Lande's gradient dynamics~\citep{lande_quantitative_1982} can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled.

### 2023-03-18 ind_files/03_BD_processes.tex
Add some references

The first term of \eqref{nD_replicator_mutator} describes changes due to faithful (non-mutational) replication, and the second describes changes due to mutation. For this reason, equation \eqref{nD_replicator_mutator} is called the \emph{replicator-mutator equation} in the evolutionary game theory literature, where the individual `types' are interpreted to be pure strategies. If in addition, each $w_i(\mathbf{x})$ is linear in $\mathbf{x}$, meaning we can write $w_i(\mathbf{x}) = \sum_{j}a_{ij}x_j$ for some set of constants $a_{ij}$, then we get the replicator-mutator equation for matrix games, and the constants $a_{ij}$ form the `payoff matrix'. As is well-known, the replicator equation (without mutation) for matrix games with $m$ pure strategies is equivalent to the generalized Lotka-Volterra equations for a community with $m-1$ species~\citep{hofbauer_evolutionary_1998}, providing the connection to community ecology.  Equation \eqref{nD_replicator_mutator} is also equivalent to Eigen's \emph{quasispecies equation} from molecular evolution if each `type' is interpreted as a genetic sequence and each $w_i(\mathbf{x})$ is a constant function\footnote{Mutational effects are often additionally assumed to act through direct `transmission probabilities' of mutating from one type to another. This means that we can write $Q_i(\mathbf{p}) = \sum_j Q_{ij}p_j$, where $Q_{ii} = 0$, and for each $j \neq i$, $Q_{ij} \geq0$ is a constant describing the probability of a $j \to i$ mutation (conditioned on the occurrence of a mutation). Substituting this into \eqref{nD_replicator_mutator} yields an equation in terms of `$Q$-matrices' or `mutation matrices' that may be more familiar to some.}\citep{page_unifying_2002}. We can now calculate how the mean of any `type level' quantity $f$, defined as $f_i$ for the $i$\textsuperscript{th} type, changes in the population (For example, if each type is a phenotype for a trait such as height, which can be assigned a numerical value, then setting $f_i = \textit{value of $i\textsuperscript{th}$ phenotype}$ gives us the mean trait value in the population). The product rule of calculus tells us that we have the relation

>>>>

The first term of \eqref{nD_replicator_mutator} describes changes due to faithful (non-mutational) replication, and the second describes changes due to mutation. For this reason, equation \eqref{nD_replicator_mutator} is called the \emph{replicator-mutator equation} in the evolutionary game theory literature, where the individual `types' are interpreted to be pure strategies~\citep{hofbauer_evolutionary_1998,page_unifying_2002,cressman_replicator_2014}. If in addition, each $w_i(\mathbf{x})$ is linear in $\mathbf{x}$, meaning we can write $w_i(\mathbf{x}) = \sum_{j}a_{ij}x_j$ for some set of constants $a_{ij}$, then we get the replicator-mutator equation for matrix games, and the constants $a_{ij}$ form the `payoff matrix'. As is well-known, the replicator equation (without mutation) for matrix games with $m$ pure strategies is equivalent to the generalized Lotka-Volterra equations for a community with $m-1$ species~\citep{hofbauer_evolutionary_1998}, providing the connection to community ecology.  Equation \eqref{nD_replicator_mutator} is also equivalent to Eigen's \emph{quasispecies equation} from molecular evolution if each `type' is interpreted as a genetic sequence and each $w_i(\mathbf{x})$ is a constant function\footnote{Mutational effects are often additionally assumed to act through direct `transmission probabilities' of mutating from one type to another. This means that we can write $Q_i(\mathbf{p}) = \sum_j Q_{ij}p_j$, where $Q_{ii} = 0$, and for each $j \neq i$, $Q_{ij} \geq0$ is a constant describing the probability of a $j \to i$ mutation (conditioned on the occurrence of a mutation). Substituting this into \eqref{nD_replicator_mutator} yields an equation in terms of `$Q$-matrices' or `mutation matrices' that may be more familiar to some.}~\citep{page_unifying_2002}. We can now calculate how the mean of any `type level' quantity $f$, defined as $f_i$ for the $i$\textsuperscript{th} type, changes in the population (For example, if each type is a phenotype for a trait such as height, which can be assigned a numerical value, then setting $f_i = \textit{value of $i\textsuperscript{th}$ phenotype}$ gives us the mean trait value in the population). The product rule of calculus tells us that we have the relation

### 2023-03-18 ind_files/04_field_eqns.tex
Add some references

for every `nice' subset $A \subset \mathcal{T}$. The Dirac mass is often written $\delta_{x_i} = \delta(y-x_i)$ as a `function' of a dummy variable $y$ which will be integrated over, since in this view the Dirac mass is a `function' that can only occur inside an integral. I will stick to the notation $\delta_{x_i}$ because it emphasizes that $\delta_{x_i}$ is supposed to represent an individual with a trait value of $x_i$ (the dummy variable $y$ can be confusing in this regard). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. Thus, if the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'

@@

which in physics notation would be a function $\nu^{(t)}(y) = \sum_{i}^{N(t)} \delta(y-x_i)$. Thus, the state space of our process is

@@

As before, I assume that the birth and death functions take the form:
\label{BD_for_kimura}
b_K(x|\psi) &=  \psi(x,t)b^{\textrm{(ind)}}(x|\psi) + \mu Q(x|\psi)\\
d_K(x|\psi) &= \psi(x,t)d^{\textrm{(ind)}}(x|\psi)
As in chapter \ref{chap_BD}, $ Q(x|\psi)$ describes birth due to mutations and $\mu \geq 0$ is a constant mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ individuals in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Note that the definition of the mean value \eqref{nD_mean} of a type level quantity $f(x)$ now becomes
\overline{f}(t) = \int\limits_{\mathcal{T}}f(x)p(x,t)dx
Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain
\label{PDE_for_kimura}
\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu Q(x|\psi)
where I have defined $w(x|\psi) \coloneqq b^{\textrm{(ind)}}(x|\psi) - d^{\textrm{(ind)}}(x|\psi)$, the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, let us define as before, the total population size and population frequency as
\label{pop_size_and_freq_for_kimura}
N_K(t) &\coloneqq \int\limits_{\mathcal{T}}\psi(x,t)dx\\
p(x,t) &\coloneqq \frac{\psi(x,t)}{N_K(t)}
The population mean fitness is:
\label{mean_fitness_for_kimura}
\overline{w}(t) = \int\limits_{\mathcal{T}}w(x|\psi)p(x,t)dx
Using the chain rule in the definition of $p(x,t)$, we can calculate:
\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\frac{\partial \psi}{\partial t}(x,t) - \frac{\psi(x,t)}{N^2_K(t)}\frac{d N_K}{dt}\\
&= \frac{1}{N_K(t)}\frac{\partial \psi}{\partial t}(x,t) - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}\frac{\partial \psi}{\partial t}(y,t)dy
Where I have used the definition of $N_K(t)$ and assumed that integrals and derivatives commute in the second line. Substituting \eqref{PDE_for_kimura}, we now obtain
\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu Q(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu Q(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} Q(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} Q(y|\psi)dy\right)
where I have used the definition of $p(x,t)$ in the second line. Using \eqref{mean_fitness_for_kimura} and rearranging the terms gives us:
\label{cts_replicator_mutator}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} Q(y|\psi)dy\right]}
This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have $m(y-z)\psi(z,t) \in  \mathcal{L}^{1}(\mathbb{R}\times\mathbb{R})$ and we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

We have: 
\int\limits_{\mathbb{R}} Q(y|\psi)dy &= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dzdy\nonumber\\
&= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dydz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)\left(\int\limits_{\mathbb{R}} m(y-z)dy\right)dz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)\int\limits_{\mathbb{R}} m(u)dudz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)dz\int\limits_{\mathbb{R}} m(u)du\nonumber\\
&= N_K(t)\int\limits_{\mathbb{R}} m(u)du\label{convolution_integral_intermediate}
where I have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} Q(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have
\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[\int\limits_{\mathbb{R}}m(x-z)\psi(z,t)dz - p(x,t)N_K(t)\right]
Substituting our definition $p(z,t) = \psi(z,t)/N_K(t)$ now yields
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\mu\left[\int\limits_{\mathbb{R}}m(x-z)p(z,t)dz - p(x,t)\right]}\label{kimura_continuum_model}
which is Kimura's continuum of alleles model.\\

@@

is the statistical covariance of the quantity $f$ with the Malthusian fitness function (Importantly, just like in the Price equation, this is an \emph{analogy} - Everything here is deterministic). The second term, which I will denote by

@@

Note that by definition, $B(y(t))$ is the statistical variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} is the canonical form of a broad class of systems called `gradient equations' or `gradient dynamics' in quantitative genetics~\citep{lande_natural_1976, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  It is also deeply related~\citep{lehtonen_price_2018, lion_theoretical_2018} to the canonical equation of adaptive dynamics~\citep{dieckmann_dynamical_1996, doebeli_adaptive_2011}. The major difference is that in the `proper' canonical equation of adaptive dynamics (as formulated in~\cite{dieckmann_dynamical_1996}), the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory~\citep{champagnat_unifying_2006}. A heuristic derivation of the canonical equation of adaptive dynamics is provided in the classic article by~\cite{dieckmann_dynamical_1996}.

>>>>

for every `nice' subset $A \subset \mathcal{T}$. The Dirac mass is often written $\delta_{x_i} = \delta(y-x_i)$ as a `function' of a dummy variable $y$ which will be integrated over, since in this view the Dirac mass is a `function' that can only occur inside an integral. Though I will indeed (kind of) use the Dirac mass as a function, I will stick to the notation $\delta_{x_i}$ because it emphasizes that $\delta_{x_i}$ is supposed to represent an individual with a trait value of $x_i$ (the dummy variable $y$ can be confusing in this regard). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. If the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'

@@

which in physics notation would be a function $\nu^{(t)}(y) = \sum_{i=1}^{N(t)} \delta(y-x_i)$. Thus, the state space of our process is

@@

As before, I assume that the birth and death functions take the form:
\begin{equation}
\label{BD_for_kimura}
\begin{aligned}
b_K(x|\psi) &=  \psi(x,t)b^{\textrm{(ind)}}(x|\psi) + \mu Q(x|\psi)\\
d_K(x|\psi) &= \psi(x,t)d^{\textrm{(ind)}}(x|\psi)
\end{aligned}
\end{equation}
As in chapter \ref{chap_BD}, $ Q(x|\psi)$ describes birth due to mutations and $\mu \geq 0$ is a constant mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ individuals in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Note that the definition of the mean value \eqref{nD_mean} of a type level quantity $f(x)$ now becomes
\begin{equation*}
\overline{f}(t) = \int\limits_{\mathcal{T}}f(x)p(x,t)dx
\end{equation*}
Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain
\begin{equation}
\label{PDE_for_kimura}
\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu Q(x|\psi)
\end{equation}
where I have defined $w(x|\psi) \coloneqq b^{\textrm{(ind)}}(x|\psi) - d^{\textrm{(ind)}}(x|\psi)$, the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, let us define as before, the total population size and population frequency as
\begin{equation}
\label{pop_size_and_freq_for_kimura}
\begin{aligned}
N_K(t) &\coloneqq \int\limits_{\mathcal{T}}\psi(x,t)dx\\
p(x,t) &\coloneqq \frac{\psi(x,t)}{N_K(t)}
\end{aligned}
\end{equation}
The population mean fitness is:
\begin{equation}
\label{mean_fitness_for_kimura}
\overline{w}(t) = \int\limits_{\mathcal{T}}w(x|\psi)p(x,t)dx
\end{equation}
Using the chain rule in the definition of $p(x,t)$, we can calculate:
\begin{align*}
\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\frac{\partial \psi}{\partial t}(x,t) - \frac{\psi(x,t)}{N^2_K(t)}\frac{d N_K}{dt}\\
&= \frac{1}{N_K(t)}\frac{\partial \psi}{\partial t}(x,t) - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}\frac{\partial \psi}{\partial t}(y,t)dy
\end{align*}
Where I have used the definition of $N_K(t)$ and assumed that integrals and derivatives commute in the second line. Substituting \eqref{PDE_for_kimura}, we now obtain
\begin{align*}
\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu Q(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu Q(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} Q(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} Q(y|\psi)dy\right)
\end{align*}
where I have used the definition of $p(x,t)$ in the second line. Using \eqref{mean_fitness_for_kimura} and rearranging the terms gives us:
\begin{equation}
\label{cts_replicator_mutator}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} Q(y|\psi)dy\right]}
\end{equation}
This is a version of the replicator-mutator equation for continuous strategy spaces when each $x$ is viewed as a strategy~\citep{cressman_replicator_2014}.

Equation \eqref{cts_replicator_mutator} also recovers Kimura's continuum-of-alleles model~\citep{kimura_stochastic_1965} when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have $m(y-z)\psi(z,t) \in  \mathcal{L}^{1}(\mathbb{R}\times\mathbb{R})$ and we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

We have: 
\begin{align}
\int\limits_{\mathbb{R}} Q(y|\psi)dy &= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dzdy\nonumber\\
&= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dydz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)\left(\int\limits_{\mathbb{R}} m(y-z)dy\right)dz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)\int\limits_{\mathbb{R}} m(u)dudz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)dz\int\limits_{\mathbb{R}} m(u)du\nonumber\\
&= N_K(t)\int\limits_{\mathbb{R}} m(u)du\label{convolution_integral_intermediate}
\end{align}
where I have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} Q(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have
\begin{equation*}
\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[\int\limits_{\mathbb{R}}m(x-z)\psi(z,t)dz - p(x,t)N_K(t)\right]
\end{equation*}
Substituting our definition $p(z,t) = \psi(z,t)/N_K(t)$ now yields
\begin{equation}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\mu\left[\int\limits_{\mathbb{R}}m(x-z)p(z,t)dz - p(x,t)\right]}\label{kimura_continuum_model}
\end{equation}
which is Kimura's continuum of alleles model~\citep{kimura_stochastic_1965,crow_introduction_1970}.

@@

is the statistical covariance of the quantity $f$ with the Malthusian fitness function. The second term, which I will denote by

@@

Note that by definition, $B(y(t))$ is the statistical variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} is the canonical form of a broad class of systems called `gradient equations' or `gradient dynamics' in quantitative genetics~\citep{lande_quantitative_1982, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  It is also deeply related~\citep{lehtonen_price_2018, lion_theoretical_2018} to the canonical equation of adaptive dynamics first formulated in~\cite{dieckmann_dynamical_1996}. The major difference is that in the `proper' canonical equation of adaptive dynamics (as formulated in~\cite{dieckmann_dynamical_1996}), the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory~\citep{champagnat_unifying_2006}. A heuristic derivation of the canonical equation of adaptive dynamics is provided in the classic article by~\cite{dieckmann_dynamical_1996}.

### 2023-03-18 ind_files/06_nontechnical_discussion.tex
Add some references

\justifying
In this thesis, we have seen how stochastic birth-death processes can be used to construct and analyze mechanistic individual-based models for the dynamics of finite populations. In doing so, we have also seen that various well-known equations of evolutionary dynamics can be recovered in the infinite population size limit. In the finite-dimensional case, the infinite population limit corresponds to the equations of population genetics and evolutionary game theory. In the infinite-dimensional case, we instead obtain the equations of quantitative genetics. In both cases, the mean value of the trait in the population changes according to an equation resembling the Price equation. My derivation highlights the natural connections between various equations of population dynamics --- For example, the same procedures that lead to the replicator-mutator equation in the case of discretely varying traits yield Kimura's continuum-of-alleles model in the quantitative case --- and extends these similarities to finite, fluctuating populations (Table \ref{table_summary}).

>>>>

\justifying
In this thesis, we have seen how stochastic birth-death processes can be used to construct and analyze mechanistic individual-based models for the dynamics of finite populations. In doing so, we have also seen that various well-known equations of evolutionary dynamics can be recovered in the infinite population size limit. In the finite-dimensional case, the infinite population limit corresponds to the equations of population genetics and evolutionary game theory. In the infinite-dimensional case, we instead obtain the equations of quantitative genetics. In both cases, the mean value of the trait in the population changes according to an equation resembling the Price equation. My derivation further highlights the natural connections between various equations of population dynamic and extends these similarities to finite, fluctuating populations (Table \ref{table_summary}).
